Honors Thesis Proposal

### 2010-09-28 honors_thesis/tec_paper/paper.tex
Renamed journal_paper and made a copy for the swarm intelligence version

@@

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms, but
in all cases the number of processors never exceeds the number of particles.
The number of iterations of PSO that the algorithms can perform is thus
inherently limited by the time it takes to evaluate the objective
function---additional processors add more particles, but do not make the
iterations go any faster.

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of twenty runs).  Increasing the swarm size from 5 to 10 has a significant
effect on the value obtained.  However, increasing the swarm size from 16 to 30
makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  Other functions show similar trends, though
often the optimal swarm size is slightly larger.  For this reason, previous
work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

In this paper we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the recommended swarm size, and simply increasing
the swarm size encounters diminishing returns.

@@

Changing the way neighbors are defined, usually called the ``topology,'' has a
significant effect on the performance of the algorithm.  In the Ring topology,
each particle has one neighbor to either side of it; in the Complete
topology\footnote{The Complete topology has often been unfortunately named Star
in the literature, which in graph theory refers to a completely different
topology.  Other names have also been used, including ``global topology'' or
gbest.  We use the graph theory term ``Complete'' in this paper.}, every
particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

@@

In distributed frameworks, sychronizing all of the machines for a round of
message passing can be expensive.  The method just described uses three rounds
of message passing for every two iterations (corresponding to steps 2, 5 and 10
in \alg{centralized}).  It is possible to perform speculative evaluation in PSO
with only one round of communication per two iterations.  However, the
methodology is tedious and is not the focus of this paper, so we defer its
description to the Appendix.

@@

@@

\subsection{Branch Statistics}

@@

We show in Table~\ref{tab:stats} with what percentage a particle takes each of
these branches for three different topologies and four different functions.
All of our statistics are from swarms of 240 particles.  Brief experimentation
showed that other swarm sizes had similar statistics.  We ran 750 iterations on
all combinations of functions and topologies except for the functions Griewank
and Rastrigin with the Complete topology.  We found that those runs frequently
converged past machine precision after 500 iterations, and that led to
erroneously high values for the probability of stagnation.  Instead we ran for
only 450 iterations on those two combinations.  All of our results were
averaged over 20 runs of the algorithm; thus the probabilities presented are
the averages of 3.6 million trials for the branch taken (2.16 million for the
two with only 450 iterations).  Table~\ref{tab:stats} contains the results.  We
have already defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock
is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.

@@

We ran experiments to compare our speculative PSO algorithm to the standard PSO
algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  We acknowledge that for benchmark functions with very fast
evaluations this may not be the best use of parallelism in PSO.  But for many
real-world applications, the objective function takes on the order of minutes
or more to evaluate; in such cases our choice of framework is reasonable.

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations.''

To use the same number of processors (and hence evaluations) for each
algorithm, speculative evaluation must use a smaller swarm size than standard
PSO.  For each of the topologies we used, a particle has three neighbors
including itself.  As shown in Table~\ref{tab:evals}, this results in $7$
speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.
The extra evaluations done in speculative evaluation would instead be particles
in regular PSO, so we compare swarms of size $p$ in speculative evaluation
with swarms of size $8p$ in regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard PSO.  But, to be fair in our comparisons, we compare to the
standard PSO algorithm using both the Random topology with two neighbors and
the Complete topology.

@@

@@

\bibliographystyle{unsrt}
\bibliography{%
../../../bib/aml/bib,%
../../../bib/nfl/bib,%
../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/sim-anneal/parallel/bib,%
../../../bib/ga/bib}

>>>>

\begin{keywords}
Parallel algorithms, optimization methods, particle swarm optimization.
\end{keywords}

@@

These previous parallel techniques distribute the computation needed by the
particles in the swarm over the available processors.  If more processors are
available, these techniques increase the number of particles in the swarm,
either by adding individual particles or by adding entire new sub-swarms.  In
almost all cases, adding additional particles produces better results in the
same amount of time.  In \fig{iters-sphere} we see an example of this on the
well-known benchmark function Sphere (20 dimensions, reporting the average of
twenty runs).  In terms of the number of iterations performed (which is
equivalent to wall-clock time if all particles are evaluated in parallel),
every time the swarm size increases, the performance improves.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

However, it can be seen from the graph that doubling the swarm size produces
only incremental benefit; performance improves, but not by very much.  The
reason for this is that there comes a point of diminishing returns with respect
to adding particles.  In \fig{evals-sphere} we show the value obtained after
50,000 function evaluations (not iterations) as a function of swarm size, again
for the function Sphere.  Increasing the swarm size from 5 to 10 has a
significant effect on the value obtained.  However, increasing the swarm size
from 16 to 30 makes the algorithm less efficient; that is, it reduces the
progress the algorithm makes per evaluation.  Other functions show similar
trends, though often the optimal swarm size is slightly larger.  For this
reason, previous work has recommended the use of a swarm size of 50 for
PSO~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in at least some
cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of function evaluations performed.  Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

In this paper we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the most efficient swarm size, and simply
adding particles results in only incremental improvement.

@@

Changing the way neighbors are defined, usually called the ``topology,'' has a
significant effect on the performance of the algorithm.  In the Ring topology,
each particle has one neighbor to either side of it; in the Complete
topology\footnote{The Complete topology has often been unfortunately named Star
in the literature, which in graph theory refers to a completely different
topology.  Other names have also been used, including ``global topology'' and
gbest.  We use the graph theory term ``Complete'' in this paper.}, every
particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

@@

In distributed frameworks, synchronizing all of the machines for a round of
message passing can be expensive.  The method just described uses three rounds
of message passing for every two iterations (corresponding to steps 2, 5 and 10
in \alg{centralized}).  It is possible to perform speculative evaluation in PSO
with only one round of communication per two iterations.  However, the
methodology is tedious and is not the focus of this paper, so we defer its
description to the Appendix.

@@

At this point it is also interesting to note a parallel between our methods and
parallel evolution
strategies~\cite{rudolph-1991-distributed-evolution-strategies}.  In evolution
strategies, a parent individual (representing a potential solution to some
objective function) produces a number of offspring by a mutation operator.  One
of the individuals is selected by a selection operator, and that individual
becomes the parent for the next
generation~\cite{beyer-2002-evolution-strategies}.  Our methods are similar,
where our mutation operator is simply the PSO motion equations and the
selection operator is either the indicator function introduced in \sec{sepso},
in the case of our original speculative algorithm, or the standard selector
operator based on fitness, in the case of this Pick Best technique.

@@

\subsection{Branch Statistics}
\label{sec:statistics}

@@

We show in Table~\ref{tab:stats} with what percentage a particle takes each of
these branches for three different topologies and four different functions.
All of our statistics are from swarms of 240 particles.  Brief experimentation
showed that other swarm sizes had similar statistics.  We ran 750 iterations on
all combinations of functions and topologies except for the functions Griewank
and Rastrigin with the Complete topology.  We found that those runs frequently
converged past machine precision after 500 iterations, and that led to
erroneously high values for the probability of stagnation.  Instead we ran for
only 450 iterations on those two combinations.  All of our results were
averaged over 20 runs of the algorithm; thus the probabilities presented are
the averages of 3.6 million trials for the branch taken (2.16 million for the
two with only 450 iterations).  Table~\ref{tab:stats} contains the results.
The definitions for all of the functions in the table are found in \sec{setup}.

@@

We ran experiments to compare our speculative PSO algorithm to the standard PSO
algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations.''

Because our interest is in exploring the most efficient use of processors, we
must hold fixed the number of processors used at each round of evaluations in
all of our experiments, giving all algorithms the same number of function
evaluations.  Thus while we label our graphs with ``Rounds of Evaluations,''
this is equivalent to the number of function evaluations used by each algorithm
(simply multiply by the number of processors to transform rounds of evaluations
to function evaluations).  We use ``Rounds of Evaluations'' to reflect
wall-clock time, as our primary interest is using extra processors in parallel
to improve performance on functions with long evaluation times.  A smaller
swarm running serial PSO would likely produce better results than we present in
the same number of function evaluations, but because the function evaluations
were not done in parallel it would take more wall-clock time.  When comparing
the performance of parallelized algorithms on long-running functions, we must
compare by wall-clock time with the same resources, not solely by the number of
function evaluations.

Using the same number of processors for each algorithm requires that our
speculative algorithms have a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has three neighbors including itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p$ evaluations for every two iterations, where $p$ is the
number of particles in the speculative swarm.  The extra evaluations done in
speculative evaluation would instead be particles in regular PSO, so we compare
swarms of size $p$ in speculative evaluation with swarms of size $8p$ in
regular PSO.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard PSO.  But, to be fair in our comparisons, we compare to the
standard PSO algorithm using both the Random topology with two neighbors and
the Complete topology (labeled PSO Random and PSO Complete in our results).

@@

In computing the branch statistics in \sec{statistics}, we also used the
Rosenbrock function.  That function is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]

@@

\bibliographystyle{unsrt}
\bibliography{%
../../../bib/aml/bib,%
../../../bib/nfl/bib,%
../../../bib/optimization/bib,%
../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/sim-anneal/parallel/bib,%
../../../bib/ga/bib}
